Portfolios of Subgraph Isomorphism Algorithms

### 2015-12-14 paper/paper.tex
Typeset algorithm names

\usepackage{times}
\usepackage{complexity}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{xcolor,colortbl}
\usepackage{tabularx}
\usepackage{siunitx}

@@

@@

The subgraph isomorphism problem is to find an adjacency-preserving injective mapping from vertices
of a small \emph{pattern} graph to vertices of a large \emph{target} graph. This \NP-complete
problem has many important practical applications, for example in computer vision
\cite{cviu11,pr15}, biochemistry \cite{Giugno:2013}, and model checking \cite{Sevegnani:2015}. There
exist various exact algorithms, which have been compared on a large suite of problem instances by
McCreesh and Prosser \cite{McCreesh:2015}. These experiments indicated that the single best
algorithm depends on the CPU time limit considered: for very small CPU time limits, VF2
\cite{Cordella:2004} is the best choice, whereas the Glasgow algorithm \cite{McCreesh:2015} has
better success rates for larger time limits.  They also showed that on an instance by instance
basis, other algorithms are often better.

@@

The simplest form of filtering is simply to propagate difference constraints (which ensure that the
matching is injective) and edge constraints (which ensure that the matching preserves pattern
edges): each time a pattern node $u\in N_p$ is matched with a target node $v\in N_t$, one removes
every candidate pair $(u',v')\in N_p\times N_t$ such that either $v'=v$ (difference constraint) or
$(u,u')$ is a pattern edge but $(v,v')$ is not a target edge (edge constraint). This simple
filtering (called \emph{Forward-Checking}) is very fast to achieve: in ${\cal O}(n_p)$ for
difference constraints, and in ${\cal O}(d_p\cdot n_t)$ for edge constraints. It is used, for
example, in McGregor's algorithm \cite{mcgregor79} and in VF2 \cite{Cordella:2004}.

R\'egin \cite{regin} introduced a stronger filtering for difference constraints, which ensures that
all pattern nodes can be matched with different target nodes, all together. This filtering (called
\emph{AllDifferent Generalized Arc Consistency}) removes more candidate pairs than when each
difference constraint is propagated separately which Forward-Checking. However, it is also more time
consuming as it is done in ${\cal O}(n_p^2\cdot n_t^2)$ time.

@@

Stronger filtering may be obtained by propagating edge constraints in a more global way, as proposed
by Larrosa and Valiente \cite{LV02}. The idea is to check for each candidate pair $(u,v)\in
N_p\times N_t$ that the number of pattern nodes adjacent to $u$ is smaller than or equal to the
number of target nodes that are both adjacent to $v$ and that may be matched with nodes adjacent to
$u$. This is done in ${\cal O}(n_p^2\cdot n_t^2)$. This idea was generalised by Solnon
\cite{Solnon:2010} where, for each candidate pair $(u,v)\in N_p\times N_t$, a redundant Local
AllDifferent (LAD) constraint ensures that each neighbour of $u$ may be matched with a different
neighbour of $v$. This is done in ${\cal O}(n_p\cdot n_t\cdot d_p^2\cdot d_t^2)$.

@@

\begin{itemize}
\item  By Battiti and Mascia \cite{battiti-mascia07}, to remove candidate pairs $(u,v)\in N_p\times
    N_t$ such that the number of paths starting from  pattern node $u$ is greater than the number of
    paths starting from  target node $v$;
\item By Audemard et al.\ \cite{Audemard:2014} to generalise the LAD constraint proposed by Solnon
    \cite{Solnon:2010} so that it ensures that a subset of pattern nodes can be matched with all
    different compatible target nodes, where compatibility is defined with respect to invariant
    properties;
\item by McCreesh and Prosser \cite{McCreesh:2015} to filter the set of candidate couples before
    starting the search, and to generate additional implied adjacency-like constraints which are
    processed during search.
\end{itemize}

@@

We measured runtimes on machines with Intel Xeon E5-2640 v2 CPUs and 64GBytes RAM, running
Scientific Linux 6.5. We used the C++ implementation of the Glasgow algorithm \cite{McCreesh:2015},
the C implementation of LAD \cite{Solnon:2010}, and the VFLib C implementation of VF2
\cite{Cordella:2004}. Software was compiled using GCC 4.9. Each problem instance was run with a
timeout of $10^8$ milliseconds (a little over a day).

@@

We have selected a set of algorithms with complementary performance:
\begin{itemize}
\item VF2 \cite{Cordella:2004}, which performs a weak filtering that it is especially fast on
    trivially satisfiable instances;
\item LAD \cite{Solnon:2010}, which combines two strong but expensive filtering techniques
    (AllDifferent Generalized Arc Consistency and LAD);
\item Glasgow \cite{McCreesh:2015}, which does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic AllDifferent propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

\noindent The Glasgow algorithm has a parameter, which controls the lengths of paths used when
reasoning about non-adjacent vertices.  In experiments reported by McCreesh and Prosser
\cite{McCreesh:2015}, the choice of paths of length 3 was used as a reasonable compromise---longer
paths lead to prohibitively expensive preprocessing on larger, denser instances. This is often not
the best choice on an instance by instance basis: sometimes path-based reasoning gives no benefit at
all, sometimes considering only paths of length 2 suffices, occasionally paths of length 4 are
helpful, and even looking at paths of length 3 is relatively expensive on some graphs.

We also introduce two new variants of LAD. The first variant is a weaker LAD filtering (called
incompleteLAD) which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check that there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
IncompleteLAD is an incomplete algorithm as it can only detect inconsistency of some instances: when
it does not detect inconsistency, the instance may either be satisfiable or not. As a counterpart of
this incompleteness, it is very fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

The second variant of LAD is called pathLAD, which combines the LAD constraints introduced by Solnon
\cite{Solnon:2010} with the exploitation of path length properties proposed by Audemard et
al.\ \cite{Audemard:2014}. The idea is to label each edge $(u,v)$ with the number of paths of length 2
between $u$ and $v$, and each node $u$ with the number of cycles of length 3 passing through $u$,
and to add the constraint that the label of a pattern node (resp.\ edge) must be smaller than or
equal to the label of its associated target node (resp.\ edge).

@@

Figure \ref{expTimeGraph} displays the evolution of the number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. IncompleteLAD is
able to solve easy unsastisfiable instances very quickly, in a few milliseconds. Hence, for time
limits lower than \SI{5}{\ms}, the best solver is incompleteLAD. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satsifiable instances.

PathLAD and Glasgow(1) outperform incompleteLAD when considering longer time limits: PathLAD is the
best solver for time limits larger than \SI{5}{\ms} and lower than \SI{40}{\ms}, and Glasgow(1) is the best solver
for time limits larger than \SI{40}{\ms} and lower than \SI{3000}{\ms}.

For CPU time limits larger than \SI{3000}{\ms}, the best solver becomes Glasgow(2).  As we
increase the CPU time limit, variants of Glasgow with longer paths (Glasgow(3) and Glasgow(4))
become better. This is what we expect, as more reasoning is expensive, but potentially increases the
reduction of the search space. In particular, the relative performance of Glasgow variants is
monotone in the time it takes to solve the instance---Glasgow(1) is best at first, but gets worse as
the time to solve the instance increases until Glasgow(2) is better. After that point, Glasgow(1) is
never better than Glasgow(2) again. The same is true for path lengths 3 and 4.

@@

Table~\ref{expClass} shows us that we cannot simply select algorithms based upon which class from
which instances are coming (unlike in SAT, where this kind of approach can be reasonable). For all
classes, there are always at least two algorithms which are the best for at least one instance of
the class. In particular, for classes 2 and 3, each algorithm is the best for at least one instance
(except Glasgow(4) for class 3).

\begin{table}[t]
\begin{center}
\begin{tabularx}{.8\textwidth}{XXrXrrrXrrrr}
\toprule
Class && VF2 && \multicolumn{3}{c}{LAD} && \multicolumn{4}{c}{\hspace*{2em}Glasgow}\\
&&&&incomplete&default&path&&1&2&3&4\\
\midrule
1 &&        0 &&       20 &        0 &        0 &&       80 &        0 &        0 &        0 \\
2 &&      201 &&       92 &      189 &      270 &&      520 &      180 &       53 &       15 \\
3 &&      112 &&     1608 &      617 &      959 &&      396 &      195 &       21 &        0 \\
4 &&      270 &&        0 &        0 &        0 &&        5 &        0 &        0 &        0 \\
5 &&      266 &&        0 &        1 &        3 &&       31 &        0 &        0 &        0 \\
6 &&       71 &&        0 &        0 &        0 &&        7 &       14 &        1 &        0 \\
7 &&      270 &&        0 &        0 &        0 &&        5 &        0 &        0 &        0 \\
8 &&        0 &&        0 &        0 &        1 &&      195 &       69 &        6 &        0 \\
9 &&       77 &&        3 &        0 &       19 &&      103 &        1 &        0 &        0 \\
10 &&       13 &&        0 &        2 &        2 &&        7 &        0 &        0 &        0 \\
11 &&        2 &&      142 &       71 &       17 &&       23 &        0 &        0 &        0 \\
12 &&        0 &&        0 &        1 &        2 &&      158 &        6 &        1 &        0 \\
\midrule
Total && 1282 && 1865 & 881 & 1273 && 1530 & 465 & 82& 15\\
\bottomrule
\end{tabularx}
\end{center}
\caption{Number of times each algorithm is best, for each class.\label{expClass}}
\end{table}

@@

Experimental results reported in Section~\ref{expComp} have shown us that incompleteLAD is very fast
(\SI{7}{\ms} on average) and is able to solve 1919 instances very quickly. Therefore, we first run
incompleteLAD: if the instance is detected to be unsatisfiable, then we do not process it further.
Otherwise, we record as features the number of candidate couples removed by incompleteLAD in absolute
terms, as a percentage, and the minimum and maximum on a per-variable basis.

VF2 is also able to solve many easy instances very quickly: among the 3806 instances which are not
solved by incompleteLAD, 1470 are solved by VF2 in less than 50 milliseconds. Therefore, after
running incompleteLAD, we run the VF2 solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.
After running incompleteLAD and VF2 for \SI{50}{\ms}, we are left with 2336 hard instances that we
consider for algorithm selection.

@@

\begin{itemize}
    \item The number of vertices and edges.
    \item The density---we expect that some kinds of filtering might be expensive and ineffective on
        dense graphs.
    \item How many loops (self-adjacent vertices) the graph contains---since loops must be mapped to
        loops, this could have a strong effect on how easy an instance is.
    \item The mean and maximum degrees, and whether or not every vertex has the same degree. (The
        degree-based invariants used by LAD and Glasgow do nothing at the top of search if every
        vertex has the same degree.)
    \item Whether or not the graph is connected.
    \item The mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective). We also consider the
        proportion of vertex pairs which are distance at least 2, 3 and 4 apart.
\end{itemize}

@@

To decide which solver to run on the remaining instances, we use LLAMA~\cite{kotthoff_llama_2013}.
LLAMA supports the most common algorithm selection approaches used in the literature. We performed a
set of preliminary experiments to determine the approach that works best in this case.

We use 10-fold cross-validation to determine the performance of the LLAMA models. The entire set of
instances was randomly partitioned into 10 subsets of approximately equal size. Of the 10 subsets, 9
were combined to form the training set for the algorithm selection models, which were evaluated on
the remaining subset. This process was repeated 10 times for all possible combinations of training
and test sets. At the end of this process, each problem instance in the original set was used
exactly once to evaluate the performance of the algorithm selection models.

LLAMA's pairwise regression approach with random forest regression gave the best performance. The
idea is very similar to the pairwise classification used by Xu et al.\ \cite{xu_satzilla_2008}. For
each pair of algorithms in our portfolio, we train a model that predicts the performance difference
between them. That is, if the first algorithm is better than the second, the difference is positive,
otherwise negative. The algorithm with the highest cumulative performance difference, i.e.\ the most
positive difference over all other algorithms, is chosen to be run.

@@

Table~\ref{tab:res} shows the performance of our algorithm selection approach, compared to two
baselines. The virtual best solver is the oracle predictor that, for each instance, chooses the best
solver from our portfolio. This is the upper bound of what an algorithm selection approach can
achieve. The single best solver is the one solver from the portfolio that has the overall best
performance across the entire set of instances, at the CPU time limit of \SI{10e8}{\ms},
i.e.\ Glasgow(2). We consider it a lower bound on performance.
We are able to solve 34 more instances than the single best solver within the timeout, with only an
additional 13 to the virtual best. In terms of average performance, we are able to close almost
\SI{70}{\percent} of the gap between the single best and the virtual best solver.

\begin{table}[ht]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 2219 & 5822809\\
single best & 1965158 & 2172 & 7787967\\
LLAMA & 610835 & 2206 & 6435301\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection performance on the reduced set of 2336
instances. MCP is the misclassification penalty; that is, the additional time
required to solve a set of instances because of choosing solvers that perform
worse than the best.}\label{tab:res} \end{table}

Table~\ref{tab:res} shows the performance of the selection model. The performance of the entire
system we propose here is shown in Table~\ref{tab:resfull}. The additional instances are solved
either by incompleteLAD or VF2 presolving, as explained above. Our system is able to close more than
\SI{50}{\percent} of the gap between single and virtual best.

\begin{table}[ht]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 5608 & 2375913\\
single best & 798660 & 5562 & 3174573\\
LLAMA & 249242 & 5595 & 2625831\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection system performance on the full set of 5725
instances.}\label{tab:resfull}
\end{table}

Figure~\ref{fig:portfolio-ecdf} shows the cumulative distribution function for
the individual solvers, the virtual best solver, and the LLAMA portfolio. The
virtual best solver clearly dominates with a significant margin to the next
best. VF2 is far below all other solvers. The portfolio does not perform well
for instances that can be solved quickly because of the overhead incurred
through feature computation. As the instances become more difficult to solve,
its performance improves.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figures/portfolio-ecdf}
\caption{Empirical cumulative distribution function of solvers, virtual best
solvers, and the LLAMA portfolio on the reduced set of 2336 instances. The
actual solvers are shown as dashed lines.}
\label{fig:portfolio-ecdf}
\end{figure}

Figure~\ref{fig:portfolio-ecdf-full} shows the ECDF of the system including
incompleteLAD and VF2 presolving on the full set of instances. Note that for
very easy instances, we are better than the virtual best solver because
incompleteLAD is not included in our portfolio. The performance on small
instances is much better than the LLAMA selector alone (cf.\
Figure~\ref{fig:portfolio-ecdf}) and the region where the portfolio performs
worse than the individual solvers is now limited to approximately $10^2$ to
$10^4$ milliseconds.

@@

The question arises of whether the success of our approach is limited to the
particular timeout we have chosen. Table~\ref{tab:resTimeouts} shows the
performance of our portfolio for timeouts ranging from $10^2$ to $10^8$. We
do not include incompleteLAD and VF2 presolving here, as these components do not
change for different timeouts.

>>>>

\usepackage{times}
\usepackage{complexity}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{xcolor,colortbl}
\usepackage{tabularx}
\usepackage{siunitx}
\usepackage{xspace}
\usepackage{scalefnt}

@@

\newcommand{\VFtwo}{$\textsc{Vf{\scalefont{0.8}2}}$\xspace}
\newcommand{\Glasgow}{$\textsc{Glasgow}$\xspace}
\newcommand{\LAD}{$\textsc{Lad}$\xspace}
\newcommand{\IncompleteLAD}{$\textsc{IncompleteLad}$\xspace}
\newcommand{\PathLAD}{$\textsc{PathLad}$\xspace}
\newcommand{\GlasgowOne}{$\textsc{Glasgow{\scalefont{0.8}1}}$\xspace}
\newcommand{\GlasgowTwo}{$\textsc{Glasgow{\scalefont{0.8}2}}$\xspace}
\newcommand{\GlasgowThree}{$\textsc{Glasgow{\scalefont{0.8}3}}$\xspace}
\newcommand{\GlasgowFour}{$\textsc{Glasgow{\scalefont{0.8}4}}$\xspace}
\newcommand{\LLAMA}{$\textsc{Llama}$\xspace}

@@

The subgraph isomorphism problem is to find an adjacency-preserving injective mapping from vertices
of a small \emph{pattern} graph to vertices of a large \emph{target} graph. This \NP-complete
problem has many important practical applications, for example in computer vision
\cite{cviu11,pr15}, biochemistry \cite{Giugno:2013}, and model checking \cite{Sevegnani:2015}. There
exist various exact algorithms, which have been compared on a large suite of problem instances by
McCreesh and Prosser \cite{McCreesh:2015}. These experiments indicated that the single best
algorithm depends on the CPU time limit considered: for very small CPU time limits, \VFtwo
\cite{Cordella:2004} is the best choice, whereas the \Glasgow algorithm \cite{McCreesh:2015} has
better success rates for larger time limits.  They also showed that on an instance by instance
basis, other algorithms are often better.

@@

The simplest form of filtering is simply to propagate difference constraints (which ensure that the
matching is injective) and edge constraints (which ensure that the matching preserves pattern
edges): each time a pattern node $u\in N_p$ is matched with a target node $v\in N_t$, one removes
every candidate pair $(u',v')\in N_p\times N_t$ such that either $v'=v$ (difference constraint) or
$(u,u')$ is a pattern edge but $(v,v')$ is not a target edge (edge constraint). This simple
filtering (called \emph{Forward-Checking}) is very fast to achieve: in ${\cal O}(n_p)$ for
difference constraints, and in ${\cal O}(d_p\cdot n_t)$ for edge constraints. It is used, for
example, in McGregor's algorithm \cite{mcgregor79} and in \VFtwo \cite{Cordella:2004}.

R\'egin \cite{regin} introduced a stronger filtering for difference constraints, which ensures that
all pattern nodes can be matched with different target nodes, all together. This filtering (called
\emph{All-Different Generalized Arc Consistency}) removes more candidate pairs than when each
difference constraint is propagated separately which Forward-Checking. However, it is also more time
consuming as it is done in ${\cal O}(n_p^2\cdot n_t^2)$ time.

@@

Stronger filtering may be obtained by propagating edge constraints in a more global way, as proposed
by Larrosa and Valiente \cite{LV02}. The idea is to check for each candidate pair $(u,v)\in
N_p\times N_t$ that the number of pattern nodes adjacent to $u$ is smaller than or equal to the
number of target nodes that are both adjacent to $v$ and that may be matched with nodes adjacent to
$u$. This is done in ${\cal O}(n_p^2\cdot n_t^2)$. This idea was generalised by Solnon
\cite{Solnon:2010}'s \LAD algorithm, where, for each candidate pair $(u,v)\in N_p\times N_t$, a redundant Local
All-Different constraint ensures that each neighbour of $u$ may be matched with a different
neighbour of $v$. This is done in ${\cal O}(n_p\cdot n_t\cdot d_p^2\cdot d_t^2)$.

@@

\begin{itemize}
\item  By Battiti and Mascia \cite{battiti-mascia07}, to remove candidate pairs $(u,v)\in N_p\times
    N_t$ such that the number of paths starting from  pattern node $u$ is greater than the number of
    paths starting from  target node $v$;
\item By Audemard et al.\ \cite{Audemard:2014} to generalise the locally all-different constraint
    proposed by Solnon \cite{Solnon:2010} so that it ensures that a subset of pattern nodes can be
    matched with all different compatible target nodes, where compatibility is defined with respect
    to invariant properties;
\item by McCreesh and Prosser \cite{McCreesh:2015} to filter the set of candidate couples before
    starting the search, and to generate additional implied adjacency-like constraints which are
    processed during search.
\end{itemize}

@@

We measured runtimes on machines with Intel Xeon E5-2640 v2 CPUs and 64GBytes RAM, running
Scientific Linux 6.5. We used the C++ implementation of the \Glasgow algorithm \cite{McCreesh:2015},
the C implementation of \LAD \cite{Solnon:2010}, and the VFLib C implementation of \VFtwo
\cite{Cordella:2004}. Software was compiled using GCC 4.9. Each problem instance was run with a
timeout of $10^8$ milliseconds (a little over a day).

@@

We have selected a set of algorithms with complementary performance:
\begin{itemize}
\item \VFtwo \cite{Cordella:2004}, which performs a weak filtering that it is especially fast on
    trivially satisfiable instances;
\item \LAD \cite{Solnon:2010}, which combines two strong but expensive filtering techniques
    (All-Different Generalized Arc Consistency and Locally All-Different);
\item \Glasgow \cite{McCreesh:2015}, which does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic All-Different propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

\noindent The \Glasgow algorithm has a parameter, which controls the lengths of paths used when
reasoning about non-adjacent vertices.  In experiments reported by McCreesh and Prosser
\cite{McCreesh:2015}, the choice of paths of length 3 was used as a reasonable compromise---longer
paths lead to prohibitively expensive preprocessing on larger, denser instances. This is often not
the best choice on an instance by instance basis: sometimes path-based reasoning gives no benefit at
all, sometimes considering only paths of length 2 suffices, occasionally paths of length 4 are
helpful, and even looking at paths of length 3 is relatively expensive on some graphs. We thus all
four choices, naming these variants \GlasgowOne through \GlasgowFour.

We also introduce two new variants of \LAD. The first, called \IncompleteLAD, is a weaker
filtering which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check that there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
\IncompleteLAD is an incomplete algorithm as it can only detect inconsistency of some instances: when
it does not detect inconsistency, the instance may either be satisfiable or not. As a counterpart of
this incompleteness, it is very fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

The second variant of \LAD is called \PathLAD, which combines the locally all-different constraints
introduced by Solnon \cite{Solnon:2010} with the exploitation of path length properties proposed by
Audemard et al.\ \cite{Audemard:2014}. The idea is to label each edge $(u,v)$ with the number of
paths of length 2 between $u$ and $v$, and each node $u$ with the number of cycles of length 3
passing through $u$, and to add the constraint that the label of a pattern node (resp.\ edge) must
be smaller than or equal to the label of its associated target node (resp.\ edge).

@@

Figure \ref{expTimeGraph} displays the evolution of the number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. \IncompleteLAD is
able to solve easy unsatisfiable instances very quickly, in a few milliseconds. Hence, for time
limits lower than \SI{5}{\ms}, the best solver is \IncompleteLAD. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satisfiable instances.

\PathLAD and \GlasgowOne outperform \IncompleteLAD when considering longer time limits: \PathLAD is the
best solver for time limits larger than \SI{5}{\ms} and lower than \SI{40}{\ms}, and \GlasgowOne is the best solver
for time limits larger than \SI{40}{\ms} and lower than \SI{3000}{\ms}.

For CPU time limits larger than \SI{3000}{\ms}, the best solver becomes \GlasgowTwo.  As we
increase the CPU time limit, variants of \Glasgow with longer paths (\GlasgowThree and \GlasgowFour)
become better. This is what we expect, as more reasoning is expensive, but potentially increases the
reduction of the search space. In particular, the relative performance of Glasgow variants is
monotone in the time it takes to solve the instance---\GlasgowOne is best at first, but gets worse as
the time to solve the instance increases until \GlasgowTwo is better. After that point, \GlasgowOne is
never better than \GlasgowTwo again. The same is true for path lengths 3 and 4.

@@

Table~\ref{expClass} shows us that we cannot simply select algorithms based upon which class from
which instances are coming (unlike in SAT, where this kind of approach can be reasonable). For all
classes, there are always at least two algorithms which are the best for at least one instance of
the class. In particular, for classes 2 and 3, each algorithm is the best for at least one instance
(except \GlasgowFour for class 3).

\begin{table}[t]
\begin{center}
\begin{tabularx}{.8\textwidth}{XXrXrrrXrrrr}
\toprule
Class && \VFtwo && \multicolumn{3}{c}{\LAD} && \multicolumn{4}{c}{\hspace*{2em}\Glasgow}\\
&&&&incomplete&default&path&&1&2&3&4\\
\midrule
1 &&        0 &&       20 &        0 &        0 &&       80 &        0 &        0 &        0 \\
2 &&      201 &&       92 &      189 &      270 &&      520 &      180 &       53 &       15 \\
3 &&      112 &&     1608 &      617 &      959 &&      396 &      195 &       21 &        0 \\
4 &&      270 &&        0 &        0 &        0 &&        5 &        0 &        0 &        0 \\
5 &&      266 &&        0 &        1 &        3 &&       31 &        0 &        0 &        0 \\
6 &&       71 &&        0 &        0 &        0 &&        7 &       14 &        1 &        0 \\
7 &&      270 &&        0 &        0 &        0 &&        5 &        0 &        0 &        0 \\
8 &&        0 &&        0 &        0 &        1 &&      195 &       69 &        6 &        0 \\
9 &&       77 &&        3 &        0 &       19 &&      103 &        1 &        0 &        0 \\
10 &&       13 &&        0 &        2 &        2 &&        7 &        0 &        0 &        0 \\
11 &&        2 &&      142 &       71 &       17 &&       23 &        0 &        0 &        0 \\
12 &&        0 &&        0 &        1 &        2 &&      158 &        6 &        1 &        0 \\
\midrule
Total && 1282 && 1865 & 881 & 1273 && 1530 & 465 & 82& 15\\
\bottomrule
\end{tabularx}
\end{center}
\caption{Number of times each algorithm is best, for each class.\label{expClass}}
\end{table}

@@

Experimental results reported in Section~\ref{expComp} have shown us that \IncompleteLAD is very fast
(\SI{7}{\ms} on average) and is able to solve 1919 instances very quickly. Therefore, we first run
\IncompleteLAD: if the instance is detected to be unsatisfiable, then we do not process it further.
Otherwise, we record as features the number of candidate couples removed by \IncompleteLAD in absolute
terms, as a percentage, and the minimum and maximum on a per-variable basis.

\VFtwo is also able to solve many easy instances very quickly: among the 3806 instances which are not
solved by \IncompleteLAD, 1470 are solved by \VFtwo in less than 50 milliseconds. Therefore, after
running \IncompleteLAD, we run the \VFtwo solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.
After running \IncompleteLAD and \VFtwo for \SI{50}{\ms}, we are left with 2336 hard instances that we
consider for algorithm selection.

@@

\begin{itemize}
    \item The number of vertices and edges.
    \item The density---we expect that some kinds of filtering might be expensive and ineffective on
        dense graphs.
    \item How many loops (self-adjacent vertices) the graph contains---since loops must be mapped to
        loops, this could have a strong effect on how easy an instance is.
    \item The mean and maximum degrees, and whether or not every vertex has the same degree. (The
        degree-based invariants used by \LAD and \Glasgow do nothing at the top of search if every
        vertex has the same degree.)
    \item Whether or not the graph is connected.
    \item The mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective). We also consider the
        proportion of vertex pairs which are distance at least 2, 3 and 4 apart.
\end{itemize}

@@

To decide which solver to run on the remaining instances, we use \LLAMA~\cite{kotthoff_llama_2013}.
\LLAMA supports the most common algorithm selection approaches used in the literature. We performed a
set of preliminary experiments to determine the approach that works best in this case.

We use 10-fold cross-validation to determine the performance of the \LLAMA models. The entire set of
instances was randomly partitioned into 10 subsets of approximately equal size. Of the 10 subsets, 9
were combined to form the training set for the algorithm selection models, which were evaluated on
the remaining subset. This process was repeated 10 times for all possible combinations of training
and test sets. At the end of this process, each problem instance in the original set was used
exactly once to evaluate the performance of the algorithm selection models.

\LLAMA's pairwise regression approach with random forest regression gave the best performance. The
idea is very similar to the pairwise classification used by Xu et al.\ \cite{xu_satzilla_2008}. For
each pair of algorithms in our portfolio, we train a model that predicts the performance difference
between them. That is, if the first algorithm is better than the second, the difference is positive,
otherwise negative. The algorithm with the highest cumulative performance difference, i.e.\ the most
positive difference over all other algorithms, is chosen to be run.

@@

Table~\ref{tab:res} shows the performance of our algorithm selection approach, compared to two
baselines. The virtual best solver is the oracle predictor that, for each instance, chooses the best
solver from our portfolio. This is the upper bound of what an algorithm selection approach can
achieve. The single best solver is the one solver from the portfolio that has the overall best
performance across the entire set of instances, at the CPU time limit of \SI{10e8}{\ms},
i.e.\ GlasgowTwo. We consider it a lower bound on performance.
We are able to solve 34 more instances than the single best solver within the timeout, with only an
additional 13 to the virtual best. In terms of average performance, we are able to close almost
\SI{70}{\percent} of the gap between the single best and the virtual best solver.

\begin{table}[ht]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 2219 & 5822809\\
single best & 1965158 & 2172 & 7787967\\
\LLAMA & 610835 & 2206 & 6435301\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection performance on the reduced set of 2336
instances. MCP is the misclassification penalty; that is, the additional time
required to solve a set of instances because of choosing solvers that perform
worse than the best.}\label{tab:res} \end{table}

Table~\ref{tab:res} shows the performance of the selection model. The performance of the entire
system we propose here is shown in Table~\ref{tab:resfull}. The additional instances are solved
either by \IncompleteLAD or \VFtwo presolving, as explained above. Our system is able to close more than
\SI{50}{\percent} of the gap between single and virtual best.

\begin{table}[ht]
    \centering\setlength{\tabcolsep}{1em}
\begin{tabular}{lrrr}
  \toprule
model & mean MCP & solved instances & mean performance\\
  \midrule
virtual best & 0 & 5608 & 2375913\\
single best & 798660 & 5562 & 3174573\\
\LLAMA & 249242 & 5595 & 2625831\\
   \bottomrule \\
\end{tabular}
\caption{Algorithm selection system performance on the full set of 5725
instances.}\label{tab:resfull}
\end{table}

Figure~\ref{fig:portfolio-ecdf} shows the cumulative distribution function for
the individual solvers, the virtual best solver, and the \LLAMA portfolio. The
virtual best solver clearly dominates with a significant margin to the next
best. \VFtwo is far below all other solvers. The portfolio does not perform well
for instances that can be solved quickly because of the overhead incurred
through feature computation. As the instances become more difficult to solve,
its performance improves.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figures/portfolio-ecdf}
\caption{Empirical cumulative distribution function of solvers, virtual best
solvers, and the \LLAMA portfolio on the reduced set of 2336 instances. The
actual solvers are shown as dashed lines.}
\label{fig:portfolio-ecdf}
\end{figure}

Figure~\ref{fig:portfolio-ecdf-full} shows the ECDF of the system including
\IncompleteLAD and \VFtwo presolving on the full set of instances. Note that for
very easy instances, we are better than the virtual best solver because
\IncompleteLAD is not included in our portfolio. The performance on small
instances is much better than the \LLAMA selector alone (cf.\
Figure~\ref{fig:portfolio-ecdf}) and the region where the portfolio performs
worse than the individual solvers is now limited to approximately $10^2$ to
$10^4$ milliseconds.

@@

The question arises of whether the success of our approach is limited to the
particular timeout we have chosen. Table~\ref{tab:resTimeouts} shows the
performance of our portfolio for timeouts ranging from $10^2$ to $10^8$. We
do not include \IncompleteLAD and \VFtwo presolving here, as these components do not
change for different timeouts.

\begin{table}[ht]
\centering
\begin{tabular}{llrrr}
  \toprule
timeout & solver & mean MCP & solved instances & mean performance\\
  \midrule
$10^2$ & virtual best & 0 & 1253 & 59\\
 & \LLAMA & 3 & 685 & 78\\
 & \Glasgow1 & 11 & 903 & 71\\
  \midrule
$10^3$ & virtual best & 0 & 1781 & 322\\
 & \LLAMA & 22 & 1284 & 569\\
 & \Glasgow1 & 126 & 1539 & 449\\
  \midrule
$10^4$ & virtual best & 0 & 1943 & 2018\\
 & \LLAMA & 186 & 1771 & 3342\\
 & \Glasgow2 & 683 & 1855 & 2701\\
  \midrule
$10^5$ & virtual best & 0 & 2045 & 14348\\
 & \LLAMA & 1346 & 2022 & 17231\\
 & \Glasgow2 & 3019 & 1988 & 17368\\
  \midrule
$10^6$ & virtual best & 0 & 2111 & 108060\\
 & \LLAMA & 10826 & 2087 & 120459\\
 & Glasgow2 & 22729 & 2063 & 130789\\
  \midrule
$10^7$ & virtual best & 0 & 2178 & 815577\\
 & \LLAMA & 91012 & 2156 & 908200\\
 & Glasgow2 & 198944 & 2129 & 1014522\\
  \midrule
$10^8$ & virtual best & 0 & 2219 & 5822809\\
 & \LLAMA & 740765 & 2204 & 6565232\\
 & Glasgow2 & 1969418 & 2172 & 7792227\\
\bottomrule
\end{tabular}
\vspace{1ex}
\caption{Algorithm selection performance for different timeouts.}\label{tab:resTimeouts}
\end{table}

The results show that while for the misclassification penalty, \LLAMA is always
better than the single best algorithm, for timeouts $10^2$ to $10^4$ it is worse
in terms of the number of solved instances and mean performance. The
misclassification penalty does not take the cost of computing features into
account, while the mean performance does---for these small timeouts, the cost
of computing the features dominates any improvement through algorithm selection.
Another explanation for this is that empirical runtime measurements are
inherently noisy, which affects the accuracy of the measurements and hence the
performance of the algorithm selection models particularly in this range.
